---
title: "Getting started with hilldiv3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hilldiv3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(hilldiv3)
```

# Overview

`hilldiv3` computes neutral, phylogenetic and functional Hill numbers from a
count table, and derives diversity partitioning, (dis)similarity, profiles,
evenness and redundancy from them.

The diversity *type* is inferred from the inputs:

* counts only &rarr; **neutral**
* counts + `tree` &rarr; **phylogenetic**
* counts + `dist` &rarr; **functional**

# Alpha diversity

```{r}
counts <- matrix(c(10, 0, 5, 2, 8, 1), nrow = 3,
                 dimnames = list(c("t1", "t2", "t3"), c("s1", "s2")))
hilldiv(counts, q = c(0, 1, 2))
```

# Partitioning and dissimilarity

```{r}
hillpart(counts)
hilldiss(counts)
```

<!-- TODO: expand with phylogenetic/functional examples, profiles (hillprof),
     evenness (hilleven), pairwise dissimilarity and ordination once the
     corresponding engine paths are implemented. -->
