---
title: "Getting started with hilldiv3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hilldiv3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

## What hilldiv3 does

`hilldiv3` measures and compares the diversity of biological communities
(OTU/ASV/MAG tables) using **Hill numbers**, a single family of metrics that
unifies richness, Shannon and Simpson diversity through one parameter, the
*diversity order* `q`. From the same framework it derives diversity
**partitioning**, **(dis)similarity**, **profiles**, **evenness** and
**redundancy**, for three flavours of diversity:

| Flavour | What it accounts for | How you ask for it |
|---|---|---|
| **Neutral** | abundances only | counts |
| **Phylogenetic** | evolutionary relatedness | counts + `tree` |
| **Functional** | trait dissimilarity | counts + `dist` |

The diversity *type* is inferred from the inputs you pass — you call the same
functions either way. You can also state it explicitly with
`type = "neutral" | "phylogenetic" | "functional"` to have it validated against
your inputs.

## The data

Every function takes a **count table** with taxa (OTUs/ASVs/MAGs) in rows and
samples in columns. A matrix is the simplest form:

```{r}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3, byrow = FALSE,
  dimnames = list(c("t1", "t2", "t3"),
                  c("s1", "s2", "s3", "s4"))
)
counts
```

Data frames, tibbles, `phyloseq` objects and `TreeSummarizedExperiment`
objects work too — see `vignette("preparing-data")` *(article on the website)*.

The package also ships a small **simulated** gut-microbiome example —
`gut_counts` (a MAG count table), `gut_tree` (a phylogeny) and `gut_traits`
(a trait table) — used throughout the website articles.

## Alpha diversity

`hilldiv()` returns Hill numbers per sample. By default it computes orders
`q = 0` (richness), `q = 1` (Shannon diversity) and `q = 2` (Simpson
diversity):

```{r}
hilldiv(counts)
```

Higher `q` down-weights rare taxa, so `qD` decreases as `q` grows unless the
sample is perfectly even. Add a tree or a distance matrix to switch flavour:

```{r}
tree <- ape::read.tree(text = "((t1:1,t2:1):1,t3:2);")
hilldiv(counts, tree = tree)              # phylogenetic
```

## Partitioning and dissimilarity

`hillpart()` splits diversity across samples into **alpha** (within-sample),
**gamma** (pooled) and **beta** (`gamma / alpha`, the number of effectively
distinct communities):

```{r}
hillpart(counts)
```

`hilldiss()` and `hillsim()` turn beta into bounded dissimilarity / similarity
metrics (Sorensen-, Jaccard-, and Unifrac-type), and `hillpair()` returns a
`dist` object of pairwise dissimilarities ready for ordination:

```{r}
hilldiss(counts, q = 1)
```

## Where to next

The website carries in-depth articles:

* **Diversity types** — neutral, phylogenetic and functional measurement.
* **Partitioning and (dis)similarity** — alpha/beta/gamma, the S/C/U/V
  metrics, and pairwise dissimilarity for ordination.
* **Profiles, evenness and redundancy** — `hillprof()`, `hilleven()`,
  `hillred()`.
* **Preparing your data** — input formats, `phyloseq`/`TreeSummarizedExperiment`,
  `tss()`, `traits2dist()` and `match_data()`.

## References

* Hill, M.O. (1973). Diversity and evenness. *Ecology*, 54, 427–432.
* Jost, L. (2007). Partitioning diversity into independent alpha and beta
  components. *Ecology*, 88, 2427–2439.
* Alberdi, A. & Gilbert, M.T.P. (2019). A guide to the application of Hill
  numbers to DNA-based diversity analyses. *Mol. Ecol. Resour.*, 19, 804–817.
</content>
