---
title: "Diversity types: neutral, phylogenetic and functional"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

"Diversity" can mean three different things, and which one you want depends on
the question you are asking:

* **Neutral** diversity counts taxa and weighs them by abundance, treating every
  taxon as equally distinct. Two communities with the same number of equally
  common taxa are equally diverse, no matter *which* taxa they are.
* **Phylogenetic** diversity also asks how *related* the taxa are. A community of
  three close cousins is less diverse than three taxa from distant branches of
  the tree of life, even if both have three taxa.
* **Functional** diversity asks how *different* the taxa are in their traits
  (size, diet, metabolism…). Three taxa that do very similar things are less
  diverse than three that do contrasting things.

`hilldiv3` measures all three from the same count table. You never switch
functions — you switch *inputs*, and the diversity type is inferred from what
you supply:

| You pass | Diversity type |
|---|---|
| counts only | **neutral** |
| counts + `tree` | **phylogenetic** |
| counts + `dist` | **functional** |

This article walks through all three with `hilldiv()`, the alpha-diversity
workhorse. The same `tree` / `dist` logic applies to every other `hill*`
function.

```{r data}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3,
  dimnames = list(c("t1", "t2", "t3"), c("s1", "s2", "s3", "s4"))
)
counts
```

## The diversity order `q`

Hill numbers are a one-parameter family. The order `q` sets how much weight
rare taxa carry:

* `q = 0` — **richness**: every taxon counts equally, abundance ignored.
* `q = 1` — **Shannon diversity**: taxa weighted by their frequency (the
  exponential of Shannon entropy).
* `q = 2` — **Simpson diversity**: dominated by common taxa (inverse Simpson
  concentration).

```{r q-orders}
hilldiv(counts, q = c(0, 1, 2))
```

All Hill numbers are in **effective number of taxa** ("how many equally-abundant
taxa would give this diversity"), so values across `q` are directly comparable.
A diversity *profile* sweeps `q` continuously — see the profiles article.

## Neutral diversity

With counts alone you get neutral diversity, which treats every taxon as equally
distinct:

```{r neutral}
hilldiv(counts)
```

## Phylogenetic diversity

Supply a phylogenetic `tree` (class `phylo`, e.g. from `ape::read.tree()`) whose
tip labels match the taxa. Diversity is then measured in units of branch length,
crediting samples that span deeper, more divergent lineages (Chao et al. 2010).

```{r phylo}
tree <- ape::read.tree(text = "((t1:1,t2:1):1,t3:2);")
hilldiv(counts, tree = tree)
```

The tip labels must cover the same taxa as the count table. `hilldiv3`
**reorders** the counts to match the tree internally — a frequent source of
silent error in older tools — and raises an informative error if the name sets
disagree. To intentionally drop taxa that are missing from one side, align them
first with `match_data()` (see the data-preparation article).

At `q = 0`, phylogenetic Hill numbers relate directly to Faith's PD.

## Functional diversity

Supply a `dist` matrix of pairwise functional distances between taxa. Diversity
then accounts for how trait-different the taxa are: a sample of three very
similar taxa is less functionally diverse than three contrasting ones (Chiu &
Chao 2014).

```{r func}
# A toy functional distance matrix over the three taxa.
fdist <- as.matrix(dist(
  data.frame(body = c(1, 0.2, 0.9), diet = c(0, 1, 1),
             row.names = c("t1", "t2", "t3"))
))
hilldiv(counts, dist = fdist)
```

### Building distances from traits

Usually you start from a **trait table** (taxa in rows, traits in columns) and
convert it with `traits2dist()`, which uses Gower distance by default so that
mixed continuous/categorical traits are handled sensibly:

```{r traits2dist}
traits <- data.frame(
  body_mass = c(1.0, 0.2, 0.9),
  diet      = factor(c("herb", "carn", "carn")),
  row.names = c("t1", "t2", "t3")
)
fdist <- traits2dist(traits)
round(fdist, 3)
```

### The distance threshold `tau`

Functional Hill numbers depend on a threshold `tau`: taxa farther apart than
`tau` are treated as fully distinct. By default `tau = max(dist)`, which makes
the measure use the full range of distances. Lowering `tau` makes more taxa
"functionally equivalent" and reduces functional diversity:

```{r tau}
hilldiv(counts, dist = fdist, tau = max(fdist))   # default
hilldiv(counts, dist = fdist, tau = max(fdist) / 2)
```

## One taxon, many flavours

Because all three share the Hill-number scale, you can compare what each lens
sees in the *same* samples:

```{r compare}
list(
  neutral       = hilldiv(counts),
  phylogenetic  = hilldiv(counts, tree = tree),
  functional    = hilldiv(counts, dist = fdist)
)
```

A sample can be neutrally diverse yet phylogenetically or functionally
redundant — quantifying exactly that gap is what `hillred()` does (see the
profiles/evenness/redundancy article).

## Contrasts with other packages

> **Advanced — skip on a first read.** This section is for users coming from
> other Hill-number tools who want to know why the numbers can differ. If you are
> just getting started, you can move on to the next article.

Other R packages compute Hill numbers, and their **phylogenetic** results can
disagree with `hilldiv3` even on identical data. These are not bugs: they are
different conventions for the *reference depth* `T` — the tree depth at which an
assemblage is read when branch lengths are converted into effective numbers of
lineages. Knowing them makes migration painless. (Neutral and functional results
match across packages; only the phylogenetic depth convention varies.)

On an **ultrametric** tree every sample shares the same depth, so all conventions
coincide. The differences below surface only on **non-ultrametric** trees (e.g.
most genome or gene trees), and grow with how uneven the abundances are.

### `hillR::hill_phylo`

Two adjustments line it up with `hilldiv3`:

* By default `hill_phylo()` returns the raw phylogenetic diversity `PD`
  (branch-length units; Faith's PD at `q = 0`), **not** an effective number of
  taxa. Pass `return_dt = TRUE` and read the `D_t` row to get the Hill number
  `PD / T`.
* It reads each sample at its *own* depth `T_j = sum(L_i a_i)`, so it corresponds
  to `reference = "sample"`, not `reference = "pool"`.

```{r hillr, eval = FALSE}
# hillR expects sites x species; hilldiv3 expects taxa x samples.
hillR::hill_phylo(t(counts), tree, q = 1, return_dt = TRUE)["D_t", ]
hilldiv(counts, tree = tree, q = 1, reference = "sample")
```

With those two changes the packages agree **exactly for `q >= 1`**. They differ
only at `q = 0`: `hillR` rebuilds `T` from a presence/absence (uniform) community
for the richness case, whereas `hilldiv3` keeps a single abundance-based `T`
across the whole `q` profile, following Chao et al. (2010). The Faith's PD
numerator is identical; only the `T` in the denominator differs.

### `hilldiv2::hilldiv`

`reference = "sample"` reproduces `hilldiv2`'s **single-vector** phylogenetic
formula exactly. `hilldiv2`'s **matrix** path is different: it folds a shared
even-pool depth *into* the diversity transform. That is fine on ultrametric trees
but breaks down on non-ultrametric ones — the branch weights no longer sum to
one, and the `q`-profile becomes discontinuous around `q = 1`. `hilldiv3`'s
`reference = "pool"` delivers comparable, common-depth values without that
pathology, because it applies the reference depth as a final rescaling rather
than inside the exponentiated sum.

| Want to match | Use in `hilldiv3` |
|---|---|
| `hillR::hill_phylo(..., return_dt = TRUE)` | `reference = "sample"` (exact for `q >= 1`) |
| `hilldiv2::hilldiv(vector, tree)` | `reference = "sample"` (exact) |
| `hilldiv2::hilldiv(matrix, tree)` | no exact match — `reference = "pool"` is the sound analogue |

In short: when results diverge, check (1) whether the other package reports `PD`
or an effective number, and (2) which reference depth it uses. `hilldiv3` makes
that choice explicit through `reference`.

## References

* Chao, A., Chiu, C.-H. & Jost, L. (2010). Phylogenetic diversity measures based
  on Hill numbers. *Phil. Trans. R. Soc. B*, 365, 3599–3609.
* Chiu, C.-H. & Chao, A. (2014). Distance-based functional diversity measures
  and their decomposition. *PLoS ONE*, 9, e100014.
* Alberdi, A. & Gilbert, M.T.P. (2019). A guide to the application of Hill
  numbers to DNA-based diversity analyses. *Mol. Ecol. Resour.*, 19, 804–817.
</content>
