---
title: "Diversity types: neutral, phylogenetic and functional"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

`hilldiv3` measures three flavours of diversity from the same count table. You
never switch functions — you switch *inputs*, and the diversity type is inferred:

| You pass | Diversity type |
|---|---|
| counts only | **neutral** |
| counts + `tree` | **phylogenetic** |
| counts + `dist` | **functional** |

This article walks through all three with `hilldiv()`, the alpha-diversity
workhorse. The same `tree` / `dist` logic applies to every other `hill*`
function.

```{r data}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3,
  dimnames = list(c("t1", "t2", "t3"), c("s1", "s2", "s3", "s4"))
)
counts
```

## The diversity order `q`

Hill numbers are a one-parameter family. The order `q` sets how much weight
rare taxa carry:

* `q = 0` — **richness**: every taxon counts equally, abundance ignored.
* `q = 1` — **Shannon diversity**: taxa weighted by their frequency (the
  exponential of Shannon entropy).
* `q = 2` — **Simpson diversity**: dominated by common taxa (inverse Simpson
  concentration).

```{r q-orders}
hilldiv(counts, q = c(0, 1, 2))
```

All Hill numbers are in **effective number of taxa** ("how many equally-abundant
taxa would give this diversity"), so values across `q` are directly comparable.
A diversity *profile* sweeps `q` continuously — see the profiles article.

## Neutral diversity

With counts alone you get neutral diversity, which treats every taxon as equally
distinct:

```{r neutral}
hilldiv(counts)
```

## Phylogenetic diversity

Supply a phylogenetic `tree` (class `phylo`, e.g. from `ape::read.tree()`) whose
tip labels match the taxa. Diversity is then measured in units of branch length,
crediting samples that span deeper, more divergent lineages (Chao et al. 2010).

```{r phylo}
tree <- ape::read.tree(text = "((t1:1,t2:1):1,t3:2);")
hilldiv(counts, tree = tree)
```

The tip labels must cover the same taxa as the count table. `hilldiv3`
**reorders** the counts to match the tree internally — a frequent source of
silent error in older tools — and raises an informative error if the name sets
disagree. To intentionally drop taxa that are missing from one side, align them
first with `match_data()` (see the data-preparation article).

At `q = 0`, phylogenetic Hill numbers relate directly to Faith's PD.

## Functional diversity

Supply a `dist` matrix of pairwise functional distances between taxa. Diversity
then accounts for how trait-different the taxa are: a sample of three very
similar taxa is less functionally diverse than three contrasting ones (Chiu &
Chao 2014).

```{r func}
# A toy functional distance matrix over the three taxa.
fdist <- as.matrix(dist(
  data.frame(body = c(1, 0.2, 0.9), diet = c(0, 1, 1),
             row.names = c("t1", "t2", "t3"))
))
hilldiv(counts, dist = fdist)
```

### Building distances from traits

Usually you start from a **trait table** (taxa in rows, traits in columns) and
convert it with `traits2dist()`, which uses Gower distance by default so that
mixed continuous/categorical traits are handled sensibly:

```{r traits2dist}
traits <- data.frame(
  body_mass = c(1.0, 0.2, 0.9),
  diet      = factor(c("herb", "carn", "carn")),
  row.names = c("t1", "t2", "t3")
)
fdist <- traits2dist(traits)
round(fdist, 3)
```

### The distance threshold `tau`

Functional Hill numbers depend on a threshold `tau`: taxa farther apart than
`tau` are treated as fully distinct. By default `tau = max(dist)`, which makes
the measure use the full range of distances. Lowering `tau` makes more taxa
"functionally equivalent" and reduces functional diversity:

```{r tau}
hilldiv(counts, dist = fdist, tau = max(fdist))   # default
hilldiv(counts, dist = fdist, tau = max(fdist) / 2)
```

## One taxon, many flavours

Because all three share the Hill-number scale, you can compare what each lens
sees in the *same* samples:

```{r compare}
list(
  neutral       = hilldiv(counts),
  phylogenetic  = hilldiv(counts, tree = tree),
  functional    = hilldiv(counts, dist = fdist)
)
```

A sample can be neutrally diverse yet phylogenetically or functionally
redundant — quantifying exactly that gap is what `hillred()` does (see the
profiles/evenness/redundancy article).

## References

* Chao, A., Chiu, C.-H. & Jost, L. (2010). Phylogenetic diversity measures based
  on Hill numbers. *Phil. Trans. R. Soc. B*, 365, 3599–3609.
* Chiu, C.-H. & Chao, A. (2014). Distance-based functional diversity measures
  and their decomposition. *PLoS ONE*, 9, e100014.
* Alberdi, A. & Gilbert, M.T.P. (2019). A guide to the application of Hill
  numbers to DNA-based diversity analyses. *Mol. Ecol. Resour.*, 19, 804–817.
</content>
