---
title: "Partitioning and (dis)similarity"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

Measuring *how different* samples are from one another is a separate question
from how diverse each one is. `hilldiv3` answers it by **partitioning** the
total (gamma) diversity of a set of samples into a within-sample (alpha) and a
between-sample (beta) component, then converting beta into bounded
(dis)similarity metrics. All of this works for neutral, phylogenetic and
functional diversity — just add a `tree` or `dist` as in the
[diversity-types article](diversity-types.html).

```{r data}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3,
  dimnames = list(c("t1", "t2", "t3"), c("s1", "s2", "s3", "s4"))
)
```

## Alpha, gamma and beta

`hillpart()` returns the three components per diversity order:

```{r hillpart}
hillpart(counts)
```

* **alpha** — the effective diversity of a typical single sample.
* **gamma** — the effective diversity of all samples pooled.
* **beta** — `gamma / alpha`, the effective number of *distinct* communities.
  It ranges from `1` (all samples identical) to `N` (all samples completely
  distinct), where `N` is the number of samples.

Beta is the multiplicative decomposition of Jost (2007): it is independent of
alpha, which is what makes it comparable across datasets with different
richness.

## Multi-scale (hierarchical) partitioning

Real sampling designs are usually **nested**: replicates within sites, sites
within regions, regions within a study. A single alpha/beta split throws that
structure away — it cannot tell you whether turnover happens *between nearby
replicates* or *between distant regions*. `hillpart()` solves this with a
`hierarchy` formula that decomposes gamma into one beta **per hierarchical
level**, so you can see at which spatial (or temporal, or organisational) scale
diversity actually turns over.

Supply the nesting as a one-sided formula, coarsest to finest, plus a
`metadata` table giving each sample's group memberships:

```{r hier-data}
set.seed(1)
tab <- matrix(rpois(12 * 8, 5), nrow = 12,
              dimnames = list(paste0("t", 1:12), paste0("s", 1:8)))
md <- data.frame(
  region = rep(c("N", "S"), each = 4),
  site   = rep(c("a", "b", "c", "d"), each = 2),
  row.names = paste0("s", 1:8)
)

hillpart(tab, hierarchy = ~ region / site, metadata = md)
```

The result is one row per `(q, scale)`, finest (`sample`) to coarsest
(`total`):

* `diversity` is the effective diversity at that scale (`A_k`): the per-sample
  alpha at `sample`, the pooled gamma at `total`, interpolating in between.
* `beta` is the turnover **gained** at that scale, `A_k / A_{k-1}` — e.g. the
  `site` beta is turnover among samples within a site, the `region` beta among
  sites within a region, the `total` beta among regions.
* `n_units` is how many units exist at that scale (8 samples → 4 sites → 2
  regions → 1 total).

The betas multiply back to the overall beta exactly:

$$
{}^qD_\gamma \;=\; {}^qD_\alpha \,\cdot\, \prod_k \beta_k .
$$

This **telescoping identity** is guaranteed by construction (and checked in the
package tests), so a hierarchical run is always consistent with the plain
`hillpart()` alpha and gamma. Arbitrarily deep hierarchies work — just chain
more terms, e.g. `~ study / region / site`. The `out = "matrix"` form returns
the same information as `alpha`, one `beta_<level>` column per transition, and
`gamma`.

### How it works (and what it assumes)

All three diversity types share one construction under **global equal sample
weighting** (every sample weighted `1/n`). Each scale's diversity is Jost's
(2007) weighted alpha,

$$
A_k \;=\;
\frac{\left(\sum_{u,i} w_i\,(m_{u,i}/C)^q\right)^{1/(1-q)}}
     {\left(\sum_{u} (n_u/n)^q\right)^{1/(1-q)}},
$$

where units $u$ pool $n_u$ of the $n$ samples, $m_{u,i}=\sum_{j\in u} c_{ij}$ is
the pooled contribution of feature $i$, $w_i$ is a per-feature *measure*, and
$C$ a grand-total normaliser. The three types differ only in those pieces:
neutral uses taxa with measure $1$ and $C=n$; **phylogenetic** uses branches
with measure $L_i$ (branch length) and $C=T_+=\sum_j T_j$ (summed tree depth);
**functional** uses taxa with measure $v_i$ (attribute contribution at threshold
`tau`) and $C=n_+$ (total count). The denominator is the numbers-equivalent of
the unit weights, which is exactly what keeps every $\beta_k \ge 1$ and
independent of $A$.

Phylogenetic and functional hierarchical partitioning simply add a `tree` or
`dist`, as everywhere else in `hilldiv3`:

```{r hier-phylo-func, eval = FALSE}
hillpart(tab, hierarchy = ~ region / site, metadata = md, tree = my_tree)
hillpart(tab, hierarchy = ~ region / site, metadata = md, dist = my_dist)
```

**Limitations and assumptions** — worth knowing before interpreting the betas:

* **Equal per-sample weighting only.** Units are weighted by their sample count
  (`n_u / n`), not by abundance or sampling effort. This is the choice that
  makes every beta independent of alpha for *all* `q` (Jost 2007); unequal
  weighting breaks that independence for `q ≠ 1` and is not offered.
* **Phylogenetic: one shared tree depth across scales.** Alpha and gamma are
  reported in effective-branch-length (PD) units and share a single mean depth
  `T_+` over *all* scales, which is what lets the betas telescope. For
  **ultrametric** trees the finest alpha equals the single-level
  phylogenetic `hillpart()` alpha exactly; for non-ultrametric trees the chain
  still telescopes, but the shared `T_+` makes the per-scale PD values
  depth-pooled rather than strictly per-sample — prefer ultrametric
  (time-calibrated) trees here.
* **Functional: one shared `tau` and `v_i` across scales.** The distance
  threshold is capped once over the whole table and the attribute contributions
  `v_i` are computed once from the fully pooled data, so they stay constant
  across scales (a prerequisite for telescoping). A per-subset `tau` is
  therefore not meaningful in a hierarchical run.
* **Hierarchical output is not (yet) wired into `hilldiss()`/`hillsim()` or
  `hillpair()`.** Those convert a single beta to bounded (dis)similarity; the
  per-level rescaling of a nested beta chain is left for future work.

This nested multiplicative decomposition across neutral, phylogenetic and
functional Hill numbers in one engine is, to our knowledge, not available in
other packages; it generalises the single-level partition above rather than
replacing it.

## From beta to (dis)similarity

Beta lives on `[1, N]`, which is awkward to compare across studies with
different sample counts. `hilldiss()` and `hillsim()` rescale it to `[0, 1]`
using the four overlap measures of Chiu et al. (2014):

| Code | Measure | Type |
|---|---|---|
| `S` | Sorensen-type | local overlap, occurrence-flavoured |
| `C` | Jaccard-type (Morisita-Horn at `q = 2`) | local overlap, abundance-flavoured |
| `U` | one-minus regional overlap | regional |
| `V` | homogeneity / Sorensen complement | regional |

```{r hilldiss}
hilldiss(counts)            # dissimilarities, one column per metric
hillsim(counts)             # similarities = 1 - dissimilarity
```

Request a single metric and order to get a plain matrix slice:

```{r single-metric}
hilldiss(counts, q = 1, metric = "C")
```

The choice of `q` matters as much as the metric: `q = 0` weights presence/
absence (turnover of rare taxa), while `q = 2` reflects shifts among dominant
taxa.

## Pairwise dissimilarity for ordination

`hilldiss()` summarises *all* samples at once. To compare samples two-by-two —
the input ordination methods expect — use `hillpair()`, which returns a `dist`
object (or a long tibble):

```{r hillpair-dist}
d <- hillpair(counts, q = 1, metric = "C")
d
```

That `dist` drops straight into ordination and clustering:

```{r ordination, fig.alt = "PCoA of pairwise Hill-number dissimilarities"}
pcoa <- cmdscale(d, k = 2)
plot(pcoa, type = "n", xlab = "PCoA 1", ylab = "PCoA 2")
text(pcoa, labels = rownames(pcoa))
```

Ask for several orders/metrics at once and you get a **named list** of `dist`
objects (e.g. `q0S`, `q1C`):

```{r hillpair-list}
ds <- hillpair(counts, q = c(0, 2), metric = c("C", "U"))
names(ds)
```

Or a tidy long table, convenient for `ggplot2` or `dplyr`:

```{r hillpair-tibble}
head(hillpair(counts, q = 1, metric = "C", out = "tibble"))
```

### Larger datasets

`hillpair()` computes every pair through the same partitioning engine, so cost
grows with the number of pairs. For many samples, run pairs in parallel by
setting `parallel = TRUE` with a `future` plan and the `furrr` package
installed:

```{r parallel, eval = FALSE}
library(future)
plan(multisession)
hillpair(counts, q = 1, metric = "C", parallel = TRUE)
```

## References

* Jost, L. (2007). Partitioning diversity into independent alpha and beta
  components. *Ecology*, 88, 2427–2439.
* Chiu, C.-H., Jost, L. & Chao, A. (2014). Phylogenetic beta diversity,
  similarity, and differentiation measures based on Hill numbers. *Ecological
  Monographs*, 84, 21–44.
</content>
