---
title: "Partitioning and (dis)similarity"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

Measuring *how different* samples are from one another is a separate question
from how diverse each one is. `hilldiv3` answers it by **partitioning** the
total (gamma) diversity of a set of samples into a within-sample (alpha) and a
between-sample (beta) component, then converting beta into bounded
(dis)similarity metrics. All of this works for neutral, phylogenetic and
functional diversity — just add a `tree` or `dist` as in the
[diversity-types article](diversity-types.html).

```{r data}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3,
  dimnames = list(c("t1", "t2", "t3"), c("s1", "s2", "s3", "s4"))
)
```

## Alpha, gamma and beta

`hillpart()` returns the three components per diversity order:

```{r hillpart}
hillpart(counts)
```

* **alpha** — the effective diversity of a typical single sample.
* **gamma** — the effective diversity of all samples pooled.
* **beta** — `gamma / alpha`, the effective number of *distinct* communities.
  It ranges from `1` (all samples identical) to `N` (all samples completely
  distinct), where `N` is the number of samples.

Beta is the multiplicative decomposition of Jost (2007): it is independent of
alpha, which is what makes it comparable across datasets with different
richness.

## From beta to (dis)similarity

Beta lives on `[1, N]`, which is awkward to compare across studies with
different sample counts. `hilldiss()` and `hillsim()` rescale it to `[0, 1]`
using the four overlap measures of Chiu et al. (2014):

| Code | Measure | Type |
|---|---|---|
| `S` | Sorensen-type | local overlap, occurrence-flavoured |
| `C` | Jaccard-type (Morisita-Horn at `q = 2`) | local overlap, abundance-flavoured |
| `U` | one-minus regional overlap | regional |
| `V` | homogeneity / Sorensen complement | regional |

```{r hilldiss}
hilldiss(counts)            # dissimilarities, one column per metric
hillsim(counts)             # similarities = 1 - dissimilarity
```

Request a single metric and order to get a plain matrix slice:

```{r single-metric}
hilldiss(counts, q = 1, metric = "C")
```

The choice of `q` matters as much as the metric: `q = 0` weights presence/
absence (turnover of rare taxa), while `q = 2` reflects shifts among dominant
taxa.

## Pairwise dissimilarity for ordination

`hilldiss()` summarises *all* samples at once. To compare samples two-by-two —
the input ordination methods expect — use `hillpair()`, which returns a `dist`
object (or a long tibble):

```{r hillpair-dist}
d <- hillpair(counts, q = 1, metric = "C")
d
```

That `dist` drops straight into ordination and clustering:

```{r ordination, fig.alt = "PCoA of pairwise Hill-number dissimilarities"}
pcoa <- cmdscale(d, k = 2)
plot(pcoa, type = "n", xlab = "PCoA 1", ylab = "PCoA 2")
text(pcoa, labels = rownames(pcoa))
```

Ask for several orders/metrics at once and you get a **named list** of `dist`
objects (e.g. `q0S`, `q1C`):

```{r hillpair-list}
ds <- hillpair(counts, q = c(0, 2), metric = c("C", "U"))
names(ds)
```

Or a tidy long table, convenient for `ggplot2` or `dplyr`:

```{r hillpair-tibble}
head(hillpair(counts, q = 1, metric = "C", out = "tibble"))
```

### Larger datasets

`hillpair()` computes every pair through the same partitioning engine, so cost
grows with the number of pairs. For many samples, run pairs in parallel by
setting `parallel = TRUE` with a `future` plan and the `furrr` package
installed:

```{r parallel, eval = FALSE}
library(future)
plan(multisession)
hillpair(counts, q = 1, metric = "C", parallel = TRUE)
```

## References

* Jost, L. (2007). Partitioning diversity into independent alpha and beta
  components. *Ecology*, 88, 2427–2439.
* Chiu, C.-H., Jost, L. & Chao, A. (2014). Phylogenetic beta diversity,
  similarity, and differentiation measures based on Hill numbers. *Ecological
  Monographs*, 84, 21–44.
</content>
