---
title: "Performance benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The benchmark suite compares `hilldiv3` with the legacy `hilldiv2` code and
the external packages `hillR`, `entropart`, `vegan` and `BAT`.

The goal is not to force every package into the same table when the operation
is not equivalent. Each package-operation row is labelled as:

| Status | Meaning |
|---|---|
| `yes` | The package has a directly comparable operation. |
| `partial` | The package has a related operation, but the output or statistical definition differs. |
| `no` | The package does not expose that operation. |
| `unavailable` | The package was not installed when the benchmark was run. |
| `failed` | The package was installed, but the benchmark call errored. |

## Reproducing the benchmark

Install the comparison packages, then run the benchmark script from the package
root:

```{r, eval = FALSE}
install.packages(c("bench", "hillR", "entropart", "vegan", "BAT"))
remotes::install_github("anttonalberdi/hilldiv2")

Sys.setenv(
  BENCH_ITERATIONS = 5,
  BENCH_N_TAXA = 200,
  BENCH_N_SAMPLES = 40
)
source("inst/benchmarks/run-benchmarks.R")
```

The script writes:

| File | Contents |
|---|---|
| `inst/benchmarks/results/performance.csv` | Support status, timing and memory table. |
| `inst/benchmarks/results/session-info.txt` | R version, platform and package versions. |

Increase `BENCH_ITERATIONS`, `BENCH_N_TAXA` or `BENCH_N_SAMPLES` for the final
publication run. Keep `session-info.txt` with the published results because
benchmark times depend on hardware, BLAS, R version and package versions.

```{r setup-results, include = FALSE}
results_path <- file.path("..", "inst", "benchmarks", "results", "performance.csv")
if (!file.exists(results_path)) {
  results_path <- file.path("inst", "benchmarks", "results", "performance.csv")
}
has_results <- file.exists(results_path)

format_seconds <- function(x) {
  ifelse(is.na(x), "", format(round(x, 4), nsmall = 4, trim = TRUE))
}

format_bytes <- function(x) {
  out <- rep("", length(x))
  ok <- !is.na(x)
  out[ok & x < 1024] <- paste0(round(x[ok & x < 1024]), " B")
  kb <- ok & x >= 1024 & x < 1024^2
  out[kb] <- paste0(round(x[kb] / 1024, 1), " KB")
  mb <- ok & x >= 1024^2 & x < 1024^3
  out[mb] <- paste0(round(x[mb] / 1024^2, 1), " MB")
  gb <- ok & x >= 1024^3
  out[gb] <- paste0(round(x[gb] / 1024^3, 2), " GB")
  out
}

if (has_results) {
  perf <- utils::read.csv(results_path, stringsAsFactors = FALSE)
  perf$time_s <- format_seconds(perf$median_seconds)
  perf$memory <- format_bytes(perf$memory_bytes)
  perf$result_size <- format_bytes(perf$result_bytes)
  perf$relative <- ifelse(
    is.na(perf$relative_to_hilldiv3),
    "",
    paste0(round(perf$relative_to_hilldiv3, 2), "x")
  )
}
```

```{r no-results, echo = FALSE, results = "asis"}
if (!has_results) {
  cat("No benchmark results have been generated yet. Run `inst/benchmarks/run-benchmarks.R` and rebuild the pkgdown site to populate this page.\n")
}
```

```{r run-summary, echo = FALSE}
if (has_results) {
  summary_table <- unique(perf[!is.na(perf$backend),
                               c("n_taxa", "n_samples", "iterations", "backend")])
  knitr::kable(summary_table, caption = "Benchmark configuration")
}
```

## Support matrix

```{r support-matrix, echo = FALSE}
if (has_results) {
  support <- perf[, c("operation", "facet", "package", "support", "note")]
  support <- support[order(support$operation, support$facet, support$package), ]
  knitr::kable(support, row.names = FALSE)
}
```

## Timing and memory

Rows marked `partial` are timed because they are useful context, but they should
not be interpreted as exact replacements for the corresponding `hilldiv3`
operation.

```{r timing-table, echo = FALSE}
if (has_results) {
  timed <- perf[perf$support %in% c("yes", "partial", "failed"), ]
  timed <- timed[, c(
    "operation", "facet", "package", "support", "time_s", "relative",
    "memory", "result_size", "error"
  )]
  timed <- timed[order(timed$operation, timed$facet, timed$package), ]
  names(timed) <- c(
    "Operation", "Facet", "Package", "Support", "Median seconds",
    "Relative to hilldiv3", "Allocated memory", "Result size", "Error"
  )
  knitr::kable(timed, row.names = FALSE)
}
```

## Notes on equivalence

`hillR` is the closest external comparator for Hill-number alpha diversity,
partitioning and pairwise comparisons across taxonomic, phylogenetic and
functional diversity.

`entropart` supports metacommunity alpha, beta and gamma diversity, including
phylogenetic and similarity-based diversity, but some outputs and assumptions
differ from the `hilldiv3` API.

`vegan` provides neutral Hill numbers through Renyi diversity and many pairwise
community dissimilarities, but it does not implement the phylogenetic,
functional or Hill-number S/C/U/V dissimilarity operations used by `hilldiv3`.

`BAT` includes biodiversity assessment tools for taxonomic, phylogenetic and
functional diversity. Its Hill-number function is a neutral alpha-diversity
comparator; its beta-diversity tools are related but not the same Hill-number
partition/dissimilarity operations.

```{r session-info, echo = FALSE, results = "asis"}
session_path <- sub("performance[.]csv$", "session-info.txt", results_path)
if (file.exists(session_path)) {
  cat("## Session info\n\n")
  cat("```text\n")
  cat(readLines(session_path), sep = "\n")
  cat("\n```\n")
}
```
