---
title: "Performance benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5,
  fig.retina = 2
)
```

The benchmark suite compares `hilldiv3` with the legacy `hilldiv2` code and
the external packages `hillR`, `entropart`, `vegan` and `BAT`.

```{r setup-results, include = FALSE}
result_candidates <- c(
  file.path("inst", "benchmarks", "results", "performance.csv"),
  file.path("..", "inst", "benchmarks", "results", "performance.csv"),
  file.path("..", "..", "inst", "benchmarks", "results", "performance.csv"),
  system.file("benchmarks", "results", "performance.csv", package = "hilldiv3")
)
result_candidates <- result_candidates[nzchar(result_candidates)]
result_candidates <- result_candidates[file.exists(result_candidates)]

has_results <- length(result_candidates) > 0
results_path <- if (has_results) result_candidates[[1]] else NA_character_
summary_path <- if (has_results) sub("performance[.]csv$", "performance-summary.csv", results_path) else NA_character_
timings_path <- if (has_results) sub("performance[.]csv$", "performance-times.csv", results_path) else NA_character_
has_summary <- has_results && file.exists(summary_path)
has_timings <- has_results && file.exists(timings_path)

format_seconds <- function(x) {
  ifelse(is.na(x), "", format(round(x, 4), nsmall = 4, trim = TRUE))
}

format_bytes <- function(x) {
  out <- rep("", length(x))
  ok <- !is.na(x)
  out[ok & x < 1024] <- paste0(round(x[ok & x < 1024]), " B")
  kb <- ok & x >= 1024 & x < 1024^2
  out[kb] <- paste0(round(x[kb] / 1024, 1), " KB")
  mb <- ok & x >= 1024^2 & x < 1024^3
  out[mb] <- paste0(round(x[mb] / 1024^2, 1), " MB")
  gb <- ok & x >= 1024^3
  out[gb] <- paste0(round(x[gb] / 1024^3, 2), " GB")
  out
}

if (has_results) {
  perf_iterations <- utils::read.csv(results_path, stringsAsFactors = FALSE)
  perf <- if (has_summary) {
    utils::read.csv(summary_path, stringsAsFactors = FALSE)
  } else {
    perf_iterations
  }
  has_iteration_results <- all(c("iteration", "seconds") %in% names(perf_iterations))
  if (!"benchmark" %in% names(perf)) {
    perf$benchmark <- perf$task
  }
  if (!"section" %in% names(perf)) {
    perf$section <- sub("[.].*$", "", perf$benchmark)
  }
  if (!"section_title" %in% names(perf)) {
    perf$section_title <- as.character(perf$operation)
  }
  perf$time_s <- format_seconds(perf$median_seconds)
  perf$memory <- format_bytes(perf$memory_bytes)
  perf$result_size <- format_bytes(perf$result_bytes)
  perf$relative <- ifelse(
    is.na(perf$relative_to_hilldiv3),
    "",
    paste0(round(perf$relative_to_hilldiv3, 2), "x")
  )
  perf$error[is.na(perf$error)] <- ""
  perf$package <- factor(
    perf$package,
    levels = c("hilldiv3", "hilldiv2", "hillR", "entropart", "vegan", "BAT")
  )

  perf$benchmark_label <- paste(perf$benchmark, perf$facet, sep = ": ")
  benchmark_levels <- unique(perf$benchmark_label[order(perf$benchmark, perf$task)])
  section_levels <- unique(perf$section[order(perf$benchmark, perf$task)])
  task_levels <- unique(perf$operation[order(perf$benchmark, perf$task)])
  perf$benchmark_label <- factor(perf$benchmark_label, levels = benchmark_levels)
  perf$section <- factor(perf$section, levels = section_levels)
  perf$operation <- factor(perf$operation, levels = task_levels)
  perf$facet <- factor(
    perf$facet,
    levels = c("neutral", "phylogenetic", "functional")
  )

  timed_plot <- perf[
    perf$support %in% c("yes", "partial") & !is.na(perf$median_seconds),
  ]

  if (has_iteration_results) {
    timing_samples <- perf_iterations
    if (!"benchmark" %in% names(timing_samples)) {
      timing_samples$benchmark <- timing_samples$task
    }
    if (!"section" %in% names(timing_samples)) {
      timing_samples$section <- sub("[.].*$", "", timing_samples$benchmark)
    }
    if (!"section_title" %in% names(timing_samples)) {
      timing_samples <- merge(
        timing_samples,
        unique(perf[, c("benchmark", "section_title")]),
        by = "benchmark",
        all.x = TRUE
      )
    }
  } else if (has_timings) {
    timing_samples <- utils::read.csv(timings_path, stringsAsFactors = FALSE)
    if (!"section" %in% names(timing_samples)) {
      timing_samples$section <- sub("[.].*$", "", timing_samples$benchmark)
    }
    if (!"section_title" %in% names(timing_samples)) {
      timing_samples <- merge(
        timing_samples,
        unique(perf[, c("benchmark", "section_title")]),
        by = "benchmark",
        all.x = TRUE
      )
    }
  } else {
    timing_samples <- timed_plot
    timing_samples$seconds <- timing_samples$median_seconds
    timing_samples$iteration <- 1L
  }

  timing_samples$package <- factor(
    timing_samples$package,
    levels = levels(perf$package)
  )
  timing_samples$facet <- factor(
    timing_samples$facet,
    levels = levels(perf$facet)
  )
  timing_samples$section <- factor(
    timing_samples$section,
    levels = levels(perf$section)
  )
  timing_samples$benchmark_label <- paste(
    timing_samples$benchmark,
    timing_samples$facet,
    sep = ": "
  )
  timing_samples$benchmark_label <- factor(
    timing_samples$benchmark_label,
    levels = levels(perf$benchmark_label)
  )
}

has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## Benchmark results

```{r no-results, echo = FALSE, results = "asis"}
if (!has_results) {
  cat("No benchmark results have been generated yet. Run `inst/benchmarks/run-benchmarks.R` and rebuild the pkgdown site to populate this page.\n")
}
```

```{r run-summary, echo = FALSE}
if (has_results) {
  summary_table <- unique(perf[!is.na(perf$backend),
                               c("n_taxa", "n_samples", "iterations", "backend")])
  knitr::kable(summary_table, caption = "Benchmark configuration")
}
```

Each benchmark is shown with three panels: **time** per iteration (left,
boxplots over the benchmark iterations), **memory** (centre, bytes allocated
per evaluation) and **comparability** (right). Within every
benchmark the facets are ordered neutral (top), phylogenetic (middle) and
functional (bottom), and the packages are named on the y axis. Numbers in the
comparability column are footnotes, explained below each figure.

```{r benchmark-section-figures, echo = FALSE, results = "asis", fig.width = 9, fig.height = 6.5}
has_patchwork <- requireNamespace("patchwork", quietly = TRUE)

if (has_results && has_ggplot && has_patchwork) {
  package_palette <- c(
    hilldiv3 = "#1b9e77", hilldiv2 = "#7570b3", hillR = "#d95f02",
    entropart = "#e7298a", vegan = "#66a61e", BAT = "#a6761d"
  )
  status_label <- c(yes = "comparable", partial = "partial",
                    failed = "failed", out_of_memory = "out of memory")
  status_fill <- c(comparable = "#a6dba0", partial = "#fdd49e",
                   failed = "#fbb4ae", `out of memory` = "#cccccc")
  facet_order <- c("neutral", "phylogenetic", "functional")
  pkg_levels <- levels(perf$package)

  base_theme <- ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      legend.position = "none",
      panel.grid.major.y = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(size = 11, face = "bold")
    )

  section_info <- unique(perf[order(perf$section), c("section", "section_title")])
  for (i in seq_len(nrow(section_info))) {
    section_id <- as.character(section_info$section[[i]])
    section_title <- as.character(section_info$section_title[[i]])

    sec <- perf[perf$section == section_id &
                  perf$support %in% c("yes", "partial", "failed", "out_of_memory"), ]
    if (nrow(sec) == 0L) next

    present <- pkg_levels[pkg_levels %in% as.character(sec$package)]
    fac <- function(x) factor(as.character(x), levels = facet_order)
    pkg <- function(x) factor(as.character(x), levels = rev(present))

    sec$facet <- fac(sec$facet)
    sec$package <- pkg(sec$package)

    caveats <- unique(sec$note[sec$support != "yes" &
                                 !is.na(sec$note) & nzchar(sec$note)])
    fn <- stats::setNames(seq_along(caveats), caveats)
    sec$footnote <- ifelse(sec$note %in% caveats, fn[sec$note], NA_integer_)
    sec$status <- factor(status_label[as.character(sec$support)],
                         levels = c("comparable", "partial", "failed", "out of memory"))
    sec$status_text <- ifelse(is.na(sec$footnote),
                              as.character(sec$status),
                              paste0(as.character(sec$status), " [", sec$footnote, "]"))

    tim <- timing_samples[timing_samples$section == section_id &
                            timing_samples$support %in% c("yes", "partial") &
                            !is.na(timing_samples$seconds) &
                            timing_samples$seconds > 0,
                          c("facet", "package", "seconds")]
    scaffold <- unique(sec[, c("facet", "package")])
    scaffold$seconds <- NA_real_
    tim <- rbind(tim, scaffold)
    tim$facet <- fac(tim$facet)
    tim$package <- pkg(tim$package)

    p_time <- ggplot2::ggplot(tim, ggplot2::aes(seconds, package, fill = package)) +
      ggplot2::geom_boxplot(outlier.alpha = 0.5, width = 0.6, na.rm = TRUE) +
      ggplot2::facet_grid(facet ~ ., scales = "free_y", space = "free_y", switch = "y") +
      ggplot2::scale_x_log10() +
      ggplot2::scale_fill_manual(values = package_palette, drop = FALSE) +
      ggplot2::labs(title = "Time", x = "Seconds / iteration (log10)", y = NULL) +
      base_theme +
      ggplot2::theme(
        strip.placement = "outside",
        strip.text.y.left = ggplot2::element_text(angle = 0, face = "bold")
      )

    p_mem <- ggplot2::ggplot(sec, ggplot2::aes(memory_bytes, package, fill = package)) +
      ggplot2::geom_col(width = 0.6, na.rm = TRUE) +
      ggplot2::facet_grid(facet ~ ., scales = "free_y", space = "free_y") +
      ggplot2::scale_x_continuous(labels = function(x) format_bytes(x), n.breaks = 3) +
      ggplot2::scale_fill_manual(values = package_palette, drop = FALSE) +
      ggplot2::labs(title = "Memory", x = "allocated", y = NULL) +
      base_theme +
      ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                     strip.text.y = ggplot2::element_blank())

    p_status <- ggplot2::ggplot(sec, ggplot2::aes(x = 1, y = package)) +
      ggplot2::geom_tile(ggplot2::aes(fill = status), color = "white", height = 0.9) +
      ggplot2::geom_text(ggplot2::aes(label = status_text), size = 2.7) +
      ggplot2::facet_grid(facet ~ ., scales = "free_y", space = "free_y") +
      ggplot2::scale_fill_manual(values = status_fill, drop = FALSE) +
      ggplot2::scale_x_continuous(limits = c(0.5, 1.5), expand = c(0, 0)) +
      ggplot2::labs(title = "Comparability", x = NULL, y = NULL) +
      base_theme +
      ggplot2::theme(axis.text = ggplot2::element_blank(),
                     panel.grid = ggplot2::element_blank(),
                     strip.text.y = ggplot2::element_blank())

    cat("\n\n## ", section_id, ": ", section_title, "\n\n", sep = "")
    print(patchwork::wrap_plots(p_time, p_mem, p_status, widths = c(3, 2, 2)))

    if (length(caveats) > 0L) {
      cat("\n\n")
      for (k in seq_along(caveats)) {
        cat("[", k, "] ", caveats[[k]], "  \n", sep = "")
      }
      cat("\n")
    }
  }
}
```

## What each tool can do

`hilldiv3` is the reference implementation. The contingency table below shows,
for every benchmark, whether each package offers a directly comparable
operation (**Yes**), a related but not identical operation (**Partial**), or no
equivalent (**No**). `n/a` means the package was not installed when the
benchmark ran. The caveats behind each *Partial* are listed as footnotes under
the corresponding figure above.

```{r comparability-table, echo = FALSE}
if (has_results) {
  sym <- c(yes = "Yes", partial = "Partial", no = "No",
           unavailable = "n/a", failed = "Failed", out_of_memory = "OOM")
  cont <- perf[order(perf$benchmark, perf$facet), ]
  cont$row <- paste0(cont$benchmark, " (", cont$facet, ")")
  pkgs <- levels(perf$package)
  rows <- unique(cont[, c("row", "operation")])
  mat <- matrix("", nrow = nrow(rows), ncol = length(pkgs),
                dimnames = list(rows$row, pkgs))
  for (j in seq_len(nrow(cont))) {
    mat[cont$row[[j]], as.character(cont$package[[j]])] <- sym[cont$support[[j]]]
  }
  contingency <- data.frame(
    Benchmark = rows$row,
    Operation = rows$operation,
    mat,
    check.names = FALSE,
    row.names = NULL
  )
  knitr::kable(contingency, row.names = FALSE)
}
```

## Full results

Memory is the number of bytes allocated during one evaluation, measured by
`bench`'s allocation profiler. Unlike the resident set size of the worker
process, it reflects the operation's own allocations rather than the size of the
loaded R session, so it is comparable across packages and independent of how
long the call runs. The complete per-operation and per-iteration tables are
written next to the benchmark script:

- [`performance-summary.csv`](https://github.com/alberdilab/hilldiv3/blob/main/inst/benchmarks/results/performance-summary.csv) — aggregated median time, allocated memory, result size and speed relative to `hilldiv3` for each package-operation.
- [`performance.csv`](https://github.com/alberdilab/hilldiv3/blob/main/inst/benchmarks/results/performance.csv) — per-iteration timings used for the boxplots.
- [`session-info.txt`](https://github.com/alberdilab/hilldiv3/blob/main/inst/benchmarks/results/session-info.txt) — R version, platform and package versions.

## Reproducing the benchmark

Install the comparison packages, then run the benchmark script from the package
root:

```{r, eval = FALSE}
install.packages(c("bench", "hillR", "entropart", "vegan", "BAT"))
remotes::install_github("anttonalberdi/hilldiv2")

Sys.setenv(
  BENCH_ITERATIONS = 10,
  BENCH_N_TAXA = 200,
  BENCH_N_SAMPLES = 50
)
source("inst/benchmarks/run-benchmarks.R")
```

The script writes:

| File | Contents |
|---|---|
| `inst/benchmarks/results/performance.csv` | Per-iteration support, timing and memory table. |
| `inst/benchmarks/results/performance-summary.csv` | Aggregated timing and memory summary table. |
| `inst/benchmarks/results/performance-times.csv` | Compatibility copy of the per-iteration table for boxplots. |
| `inst/benchmarks/results/session-info.txt` | R version, platform and package versions. |

Adjust `BENCH_ITERATIONS`, `BENCH_N_TAXA` or `BENCH_N_SAMPLES` to scale the run.
Each call runs in a forked worker capped at `BENCH_MEMORY_LIMIT_GB` (default 10);
a call that exceeds it is stopped and reported as `out_of_memory`.
Keep `session-info.txt` with the published results because
benchmark times depend on hardware, BLAS, R version and package versions.

## Notes on equivalence

`hillR` is the closest external comparator for Hill-number alpha diversity,
partitioning and pairwise comparisons across taxonomic, phylogenetic and
functional diversity.

`entropart` supports metacommunity alpha, beta and gamma diversity, including
phylogenetic and similarity-based diversity, but some outputs and assumptions
differ from the `hilldiv3` API.

`vegan` provides neutral Hill numbers through Renyi diversity and many pairwise
community dissimilarities, but it does not implement the phylogenetic,
functional or Hill-number S/C/U/V dissimilarity operations used by `hilldiv3`.

`BAT` includes biodiversity assessment tools for taxonomic, phylogenetic and
functional diversity. Its Hill-number function is a neutral alpha-diversity
comparator; its beta-diversity tools are related but not the same Hill-number
partition/dissimilarity operations.

```{r session-info, echo = FALSE, results = "asis"}
if (has_results) {
  session_path <- sub("performance[.]csv$", "session-info.txt", results_path)
}
if (has_results && file.exists(session_path)) {
  cat("## Session info\n\n")
  cat("```text\n")
  cat(readLines(session_path), sep = "\n")
  cat("\n```\n")
}
```
