---
title: "Performance benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5,
  fig.retina = 2
)
```

The benchmark suite compares `hilldiv3` with the legacy `hilldiv2` code and
the external packages `hillR`, `entropart`, `vegan` and `BAT`.

```{r setup-results, include = FALSE}
result_candidates <- c(
  file.path("inst", "benchmarks", "results", "performance.csv"),
  file.path("..", "inst", "benchmarks", "results", "performance.csv"),
  file.path("..", "..", "inst", "benchmarks", "results", "performance.csv"),
  system.file("benchmarks", "results", "performance.csv", package = "hilldiv3")
)
result_candidates <- result_candidates[nzchar(result_candidates)]
result_candidates <- result_candidates[file.exists(result_candidates)]

has_results <- length(result_candidates) > 0
results_path <- if (has_results) result_candidates[[1]] else NA_character_
summary_path <- if (has_results) sub("performance[.]csv$", "performance-summary.csv", results_path) else NA_character_
timings_path <- if (has_results) sub("performance[.]csv$", "performance-times.csv", results_path) else NA_character_
has_summary <- has_results && file.exists(summary_path)
has_timings <- has_results && file.exists(timings_path)

format_seconds <- function(x) {
  ifelse(is.na(x), "", format(round(x, 4), nsmall = 4, trim = TRUE))
}

format_bytes <- function(x) {
  out <- rep("", length(x))
  ok <- !is.na(x)
  out[ok & x < 1024] <- paste0(round(x[ok & x < 1024]), " B")
  kb <- ok & x >= 1024 & x < 1024^2
  out[kb] <- paste0(round(x[kb] / 1024, 1), " KB")
  mb <- ok & x >= 1024^2 & x < 1024^3
  out[mb] <- paste0(round(x[mb] / 1024^2, 1), " MB")
  gb <- ok & x >= 1024^3
  out[gb] <- paste0(round(x[gb] / 1024^3, 2), " GB")
  out
}

if (has_results) {
  perf_iterations <- utils::read.csv(results_path, stringsAsFactors = FALSE)
  perf <- if (has_summary) {
    utils::read.csv(summary_path, stringsAsFactors = FALSE)
  } else {
    perf_iterations
  }
  has_iteration_results <- all(c("iteration", "seconds") %in% names(perf_iterations))
  if (!"benchmark" %in% names(perf)) {
    perf$benchmark <- perf$task
  }
  if (!"section" %in% names(perf)) {
    perf$section <- sub("[.].*$", "", perf$benchmark)
  }
  if (!"section_title" %in% names(perf)) {
    perf$section_title <- as.character(perf$operation)
  }
  perf$time_s <- format_seconds(perf$median_seconds)
  perf$memory <- format_bytes(perf$memory_bytes)
  perf$result_size <- format_bytes(perf$result_bytes)
  perf$relative <- ifelse(
    is.na(perf$relative_to_hilldiv3),
    "",
    paste0(round(perf$relative_to_hilldiv3, 2), "x")
  )
  perf$error[is.na(perf$error)] <- ""
  perf$package <- factor(
    perf$package,
    levels = c("hilldiv3", "hilldiv2", "hillR", "entropart", "vegan", "BAT")
  )

  perf$benchmark_label <- paste(perf$benchmark, perf$facet, sep = ": ")
  benchmark_levels <- unique(perf$benchmark_label[order(perf$benchmark, perf$task)])
  section_levels <- unique(perf$section[order(perf$benchmark, perf$task)])
  task_levels <- unique(perf$operation[order(perf$benchmark, perf$task)])
  perf$benchmark_label <- factor(perf$benchmark_label, levels = benchmark_levels)
  perf$section <- factor(perf$section, levels = section_levels)
  perf$operation <- factor(perf$operation, levels = task_levels)
  perf$facet <- factor(
    perf$facet,
    levels = c("neutral", "phylogenetic", "functional")
  )

  timed_plot <- perf[
    perf$support %in% c("yes", "partial") & !is.na(perf$median_seconds),
  ]

  if (has_iteration_results) {
    timing_samples <- perf_iterations
    if (!"benchmark" %in% names(timing_samples)) {
      timing_samples$benchmark <- timing_samples$task
    }
    if (!"section" %in% names(timing_samples)) {
      timing_samples$section <- sub("[.].*$", "", timing_samples$benchmark)
    }
    if (!"section_title" %in% names(timing_samples)) {
      timing_samples <- merge(
        timing_samples,
        unique(perf[, c("benchmark", "section_title")]),
        by = "benchmark",
        all.x = TRUE
      )
    }
  } else if (has_timings) {
    timing_samples <- utils::read.csv(timings_path, stringsAsFactors = FALSE)
    if (!"section" %in% names(timing_samples)) {
      timing_samples$section <- sub("[.].*$", "", timing_samples$benchmark)
    }
    if (!"section_title" %in% names(timing_samples)) {
      timing_samples <- merge(
        timing_samples,
        unique(perf[, c("benchmark", "section_title")]),
        by = "benchmark",
        all.x = TRUE
      )
    }
  } else {
    timing_samples <- timed_plot
    timing_samples$seconds <- timing_samples$median_seconds
    timing_samples$iteration <- 1L
  }

  timing_samples$package <- factor(
    timing_samples$package,
    levels = levels(perf$package)
  )
  timing_samples$facet <- factor(
    timing_samples$facet,
    levels = levels(perf$facet)
  )
  timing_samples$section <- factor(
    timing_samples$section,
    levels = levels(perf$section)
  )
  timing_samples$benchmark_label <- paste(
    timing_samples$benchmark,
    timing_samples$facet,
    sep = ": "
  )
  timing_samples$benchmark_label <- factor(
    timing_samples$benchmark_label,
    levels = levels(perf$benchmark_label)
  )
}

has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## Benchmark results

```{r no-results, echo = FALSE, results = "asis"}
if (!has_results) {
  cat("No benchmark results have been generated yet. Run `inst/benchmarks/run-benchmarks.R` and rebuild the pkgdown site to populate this page.\n")
}
```

```{r run-summary, echo = FALSE}
if (has_results) {
  summary_table <- unique(perf[!is.na(perf$backend),
                               c("n_taxa", "n_samples", "iterations", "backend")])
  knitr::kable(summary_table, caption = "Benchmark configuration")
}
```

```{r benchmark-section-boxplots, echo = FALSE, results = "asis", fig.height = 5.2}
if (has_results && has_ggplot && nrow(timing_samples) > 0) {
  section_info <- unique(perf[order(perf$section), c("section", "section_title")])
  for (i in seq_len(nrow(section_info))) {
    section_id <- as.character(section_info$section[[i]])
    section_title <- as.character(section_info$section_title[[i]])
    section_data <- timing_samples[
      timing_samples$section == section_id &
        timing_samples$support %in% c("yes", "partial") &
        !is.na(timing_samples$seconds) &
        timing_samples$seconds > 0,
    ]
    if (nrow(section_data) == 0L) {
      next
    }

    cat("\n\n## ", section_id, ": ", section_title, "\n\n", sep = "")
    print(
      ggplot2::ggplot(
        section_data,
        ggplot2::aes(
          x = seconds,
          y = benchmark_label,
          fill = package
        )
      ) +
        ggplot2::geom_boxplot(
          alpha = 0.76,
          outlier.alpha = 0.55,
          width = 0.68,
          position = ggplot2::position_dodge2(width = 0.78, preserve = "single")
        ) +
        ggplot2::scale_x_log10() +
        ggplot2::scale_fill_manual(
          values = c(
            hilldiv3 = "#1b9e77",
            hilldiv2 = "#7570b3",
            hillR = "#d95f02",
            entropart = "#e7298a",
            vegan = "#66a61e",
            BAT = "#a6761d"
          ),
          drop = FALSE
        ) +
        ggplot2::labs(
          x = "Seconds per iteration, log10 scale",
          y = NULL,
          fill = "Package"
        ) +
        ggplot2::theme_minimal(base_size = 12) +
        ggplot2::theme(
          legend.position = "top",
          panel.grid.major.y = ggplot2::element_blank()
        )
    )
  }
}
```

## Support and Detail

The goal is not to force every package into the same table when the operation
is not equivalent. Each package-operation row is labelled as:

| Status | Meaning |
|---|---|
| `yes` | The package has a directly comparable operation. |
| `partial` | The package has a related operation, but the output or statistical definition differs. |
| `no` | The package does not expose that operation. |
| `unavailable` | The package was not installed when the benchmark was run. |
| `failed` | The package was installed, but the benchmark call errored. |
| `out_of_memory` | The benchmark call exceeded the memory limit and was stopped. |

### Support matrix

```{r support-matrix, echo = FALSE}
if (has_results) {
  support <- perf[, c("benchmark", "operation", "facet", "package", "support", "note")]
  support <- support[order(support$benchmark, support$facet, support$package), ]
  names(support) <- c("Benchmark", "Operation", "Facet", "Package", "Support", "Note")
  knitr::kable(support, row.names = FALSE)
}
```

### Timing and memory

Rows marked `partial` are timed because they are useful context, but they should
not be interpreted as exact replacements for the corresponding `hilldiv3`
operation. Peak memory is the maximum resident set size (RSS) of the forked
worker process during the run, so it includes the base R session and loaded
packages, not only the operation's own allocations.

```{r timing-table, echo = FALSE}
if (has_results) {
  timed <- perf[perf$support %in% c("yes", "partial", "failed", "out_of_memory"), ]
  timed <- timed[, c(
    "benchmark", "operation", "facet", "package", "support", "time_s", "relative",
    "memory", "result_size", "error"
  )]
  timed <- timed[order(timed$benchmark, timed$facet, timed$package), ]
  names(timed) <- c(
    "Benchmark", "Operation", "Facet", "Package", "Support", "Median seconds",
    "Relative to hilldiv3", "Peak memory (RSS)", "Result size", "Error"
  )
  knitr::kable(timed, row.names = FALSE)
}
```

## Reproducing the benchmark

Install the comparison packages, then run the benchmark script from the package
root:

```{r, eval = FALSE}
install.packages(c("bench", "hillR", "entropart", "vegan", "BAT"))
remotes::install_github("anttonalberdi/hilldiv2")

Sys.setenv(
  BENCH_ITERATIONS = 10,
  BENCH_N_TAXA = 200,
  BENCH_N_SAMPLES = 50
)
source("inst/benchmarks/run-benchmarks.R")
```

The script writes:

| File | Contents |
|---|---|
| `inst/benchmarks/results/performance.csv` | Per-iteration support, timing and memory table. |
| `inst/benchmarks/results/performance-summary.csv` | Aggregated timing and memory summary table. |
| `inst/benchmarks/results/performance-times.csv` | Compatibility copy of the per-iteration table for boxplots. |
| `inst/benchmarks/results/session-info.txt` | R version, platform and package versions. |

Adjust `BENCH_ITERATIONS`, `BENCH_N_TAXA` or `BENCH_N_SAMPLES` to scale the run.
Each call runs in a forked worker capped at `BENCH_MEMORY_LIMIT_GB` (default 10);
a call that exceeds it is stopped and reported as `out_of_memory`.
Keep `session-info.txt` with the published results because
benchmark times depend on hardware, BLAS, R version and package versions.

## Notes on equivalence

`hillR` is the closest external comparator for Hill-number alpha diversity,
partitioning and pairwise comparisons across taxonomic, phylogenetic and
functional diversity.

`entropart` supports metacommunity alpha, beta and gamma diversity, including
phylogenetic and similarity-based diversity, but some outputs and assumptions
differ from the `hilldiv3` API.

`vegan` provides neutral Hill numbers through Renyi diversity and many pairwise
community dissimilarities, but it does not implement the phylogenetic,
functional or Hill-number S/C/U/V dissimilarity operations used by `hilldiv3`.

`BAT` includes biodiversity assessment tools for taxonomic, phylogenetic and
functional diversity. Its Hill-number function is a neutral alpha-diversity
comparator; its beta-diversity tools are related but not the same Hill-number
partition/dissimilarity operations.

```{r session-info, echo = FALSE, results = "asis"}
if (has_results) {
  session_path <- sub("performance[.]csv$", "session-info.txt", results_path)
}
if (has_results && file.exists(session_path)) {
  cat("## Session info\n\n")
  cat("```text\n")
  cat(readLines(session_path), sep = "\n")
  cat("\n```\n")
}
```
