---
title: "Preparing your data"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

Every `hilldiv3` function accepts the same kinds of input and shares one
validation/alignment layer, so the data-wrangling you do once carries through
the whole workflow. This article covers the accepted input formats, how counts
are matched to a tree or distance matrix, and the helpers `tss()`,
`traits2dist()` and `match_data()`.

## The count table

The core input is a **count table** with taxa (OTUs/ASVs/MAGs) in rows and
samples in columns. Row names are taxa, column names are samples.

```{r counts}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3,
  dimnames = list(c("t1", "t2", "t3"), c("s1", "s2", "s3", "s4"))
)
counts
```

## Accepted input objects

You can pass any of the following — they are all coerced internally to the same
representation, so you call the functions the same way:

* a **numeric matrix** (taxa × samples);
* a **data frame** or **tibble** of counts with taxa as row names;
* a **numeric vector** for a single sample (names are the taxa);
* a [`phyloseq`](https://joey711.github.io/phyloseq/) object — its OTU table is
  read, transposed if needed, and an embedded phylogeny is picked up
  automatically;
* a
  [`TreeSummarizedExperiment`](https://bioconductor.org/packages/TreeSummarizedExperiment/)
  — its assay is used as counts and its `rowTree` as the phylogeny.

```{r vector}
# A single sample as a named vector.
hilldiv(c(t1 = 10, t2 = 2, t3 = 3))
```

When you pass a `phyloseq` or `TreeSummarizedExperiment` that already carries a
tree, phylogenetic diversity works without supplying `tree` yourself:

```{r phyloseq, eval = requireNamespace("phyloseq", quietly = TRUE)}
library(phyloseq)
ps <- phyloseq(
  otu_table(counts, taxa_are_rows = TRUE),
  phy_tree(ape::read.tree(text = "((t1:1,t2:1):1,t3:2);"))
)
hilldiv(ps)        # neutral + phylogenetic, tree taken from the object
```

## Normalisation with `tss()`

Hill numbers are computed from relative abundances, and the `hill*` functions
normalise internally — you do **not** need to pre-normalise for them. `tss()`
(total-sum scaling) is exposed for when you want normalised values yourself, or
to inspect compositions:

```{r tss}
tss(counts)             # each column sums to 1
```

It accepts a vector or a matrix and safely maps all-zero columns to zero rather
than `NaN`.

## Functional distances with `traits2dist()`

Functional diversity needs a pairwise **distance matrix** over taxa. Build one
from a trait table with `traits2dist()`, which defaults to Gower distance so
mixed continuous and categorical traits are handled correctly. Constant traits
(which carry no information) are dropped automatically.

```{r traits2dist}
traits <- data.frame(
  body_mass = c(1.0, 0.2, 0.9),
  diet      = factor(c("herb", "carn", "carn")),
  row.names = c("t1", "t2", "t3")
)
fdist <- traits2dist(traits)
round(fdist, 3)

hilldiv(counts, dist = fdist)
```

## Matching counts to a tree or distances

Phylogenetic and functional analyses require the taxa in the count table to
line up with the tree tips or distance-matrix names. `hilldiv3` enforces this
in one place:

* it checks the two name sets with `setequal()` (order does not matter);
* it **reorders** the counts to match the reference — eliminating the silent
  misalignment that plagued earlier tools;
* if the sets disagree, it stops with a message listing the offending taxa.

```{r mismatch, error = TRUE}
bad_tree <- ape::read.tree(text = "((t1:1,t2:1):1,t9:2);")
hilldiv(counts, tree = bad_tree)
```

### Intentionally dropping taxa with `match_data()`

Real datasets rarely match perfectly — a reference tree may include taxa you
never observed, or your table may carry taxa absent from the tree. `match_data()`
restricts and reorders the count table to the shared taxa, reporting what it
dropped, so you can curate alignment *before* analysis:

```{r match_data}
tree <- ape::read.tree(text = "((t1:1,t2:1):1,t4:2);")   # has t4, lacks t3
matched <- match_data(counts, tree = tree)
matched
```

Feed the aligned table back in — now the name sets agree and the analysis runs:

```{r match_data_use}
tree3 <- ape::read.tree(text = "(t1:1,t2:1);")
hilldiv(match_data(counts, tree = tree3), tree = tree3)
```

`match_data()` works the same way with a `dist` argument for functional data.

## A typical workflow

Putting it together:

```{r workflow}
# 1. Counts + traits in hand.
traits <- data.frame(
  body_mass = c(1.0, 0.2, 0.9),
  diet      = factor(c("herb", "carn", "carn")),
  row.names = c("t1", "t2", "t3")
)

# 2. Derive a functional distance matrix.
fdist <- traits2dist(traits)

# 3. Align counts to it (here they already match).
counts2 <- match_data(counts, dist = fdist)

# 4. Measure and compare.
hilldiv(counts2, dist = fdist)
hilldiss(counts2, dist = fdist, q = 1)
```
</content>
