---
title: "Profiles, evenness and redundancy"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(hilldiv3)
```

Beyond a single diversity number, `hilldiv3` offers three diagnostics that
describe the *shape* of a community: **profiles** (how diversity changes with
the order `q`), **evenness** (how equitably abundance is spread), and
**redundancy** (how much phylogenetic/functional structure is duplicated).

```{r data}
counts <- matrix(
  c(10, 0, 5,
     2, 8, 1,
     3, 4, 0,
     6, 2, 7),
  nrow = 3,
  dimnames = list(c("t1", "t2", "t3"), c("s1", "s2", "s3", "s4"))
)
tree <- ape::read.tree(text = "((t1:1,t2:1):1,t3:2);")
```

## Diversity profiles

A single `q` can mislead: two samples can have identical richness but very
different Shannon or Simpson diversity, and their ranking can even flip as `q`
increases. A **profile** plots Hill numbers across a continuous sweep of `q`,
making the full picture visible. `hillprof()` evaluates a fine grid of orders
(0 to 3 by default):

```{r hillprof}
prof <- hillprof(counts)
head(prof)
```

The result is a tidy long-format object with a ready-made plot method — one
declining curve per sample:

```{r profile-plot, fig.alt = "Hill-number diversity profiles, one curve per sample"}
plot(prof)
```

How to read it:

* The **height** at `q = 0` is richness; the height as `q → ∞` reflects
  dominance.
* A **flat** curve means an even community (diversity barely changes with `q`);
  a **steep** curve means a few taxa dominate.
* If two curves **cross**, neither sample is unambiguously more diverse — the
  answer depends on how much you weight rare taxa.

Profiles work for phylogenetic and functional diversity too, and you can grab a
matrix instead of the tidy form:

```{r profile-variants}
phylo_prof <- hillprof(counts, tree = tree)
mat <- hillprof(counts, out = "matrix")
mat[1:3, ]
```

## Evenness

Evenness asks how *equitably* abundance is distributed, independent of richness.
`hilleven()` expresses it as the ratio of diversity of order `q` to richness
(`qD / 0D`), which is bounded in `[0, 1]`: `1` is perfectly even, lower values
mean stronger dominance.

```{r hilleven}
hilleven(counts)
```

This is exactly the *normalised height* of the diversity profile — a single
sample's profile flatness summarised as a number. Reading down a column tells
you how dominated each sample is; `s2` here is the least even.

## Redundancy

Neutral diversity counts taxa; phylogenetic and functional diversity count
*distinct lineages or trait combinations*. When taxa are closely related or
functionally similar, much of the neutral diversity is **redundant** — losing a
taxon costs little unique phylogenetic/functional diversity.

`hillred()` quantifies this by fitting the saturating relationship between
neutral diversity (x) and phylogenetic/functional diversity (y) across samples,
`y = -a · 2^(-x / b) + c`, and summarising redundancy as `1 - b / max(x)`. It
therefore needs a `tree` or `dist`, and several samples to fit the curve:

```{r hillred}
hillred(counts, tree = tree)
```

The returned matrix has the `redundancy` summary plus the fitted `a`, `b`, `c`
coefficients per order. Values near `1` mean the assemblage is highly
redundant (taxa are phylogenetically/functionally interchangeable); values near
`0` mean each taxon contributes largely unique diversity.

> **Note.** Redundancy is a curve fit across samples and can fail to converge
> for some orders on small or degenerate datasets, in which case that row is
> `NA` and a warning explains why. Fit it on real datasets with enough samples
> spanning a range of neutral diversity.

## References

* Hill, M.O. (1973). Diversity and evenness. *Ecology*, 54, 427–432.
* Chao, A., Chiu, C.-H. & Jost, L. (2010). Phylogenetic diversity measures based
  on Hill numbers. *Phil. Trans. R. Soc. B*, 365, 3599–3609.
* Alberdi, A. & Gilbert, M.T.P. (2019). A guide to the application of Hill
  numbers to DNA-based diversity analyses. *Mol. Ecol. Resour.*, 19, 804–817.
</content>
