---
title: "Use case: dietary niches of insectivorous bats"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.6,
  dpi = 150,
  message = FALSE
)
```

```{r setup}
library(hilldiv3)
library(ggplot2)

theme_set(theme_bw(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  legend.position = "top"))

# hilldiv3 results are long-format data.frame subclasses; drop the class so
# base `$<-` / `rbind` (and ggplot2) work without coercion.
as_df <- function(x) { class(x) <- "data.frame"; x }
pal <- c(forest = "#2E7D32", farmland = "#C9A227")
```

Dietary metabarcoding turns faecal or guano samples into a count table of prey
amplicon sequence variants (ASVs), and a question that immediately follows is
*how diverse, and how different, are the diets of the animals we sampled?*
Because the prey of a predator are an arbitrary slice of the tree of life,
counting ASVs alone can mislead: a bat that takes twenty species of moth has a
taxon-rich but phylogenetically narrow diet, whereas a bat that takes a beetle,
a fly and a true bug eats fewer taxa spread across far more of the insect
phylogeny. `hilldiv3` makes both readings available from the same call, and
lets us ask *at which level of a nested sampling design* dietary turnover
actually occurs.

## The data

We use a simulated guano-metabarcoding design: **50 insect-prey ASVs** detected
across **30 individual bats**, nested as 5 individuals × 3 roosts × 2
contrasting foraging habitats (forest vs. farmland). Prey ASVs belong to four
insect orders, and their phylogeny has shallow within-order clades on a deep
order-level backbone. Forest bats are specialists, concentrating on a single
dominant moth (Lepidoptera) clade; farmland bats are generalists spread evenly
across Diptera, Coleoptera and Hemiptera.

The full simulation ships with the package (and is reproducible); we simply
source it here to obtain a count table (`bat_counts`), a prey phylogeny
(`bat_tree`) and a sample-design table (`bat_metadata`).

```{r data}
sim <- source(system.file("manuscript", "data-bat-diet.R",
                          package = "hilldiv3"))$value
bat_counts   <- sim$counts
bat_tree     <- sim$tree
bat_metadata <- sim$metadata

dim(bat_counts)                 # 50 prey ASVs x 30 bats
range(colSums(bat_counts > 0))  # per-sample prey richness
head(bat_metadata)
```

```{r helpers}
hab <- function(s) bat_metadata$habitat[match(s, bat_metadata$bat)]
```

## Neutral and phylogenetic diversity from one interface

`hilldiv()` returns Hill numbers per sample at diversity orders `q = 0`
(richness), `q = 1` (Shannon) and `q = 2` (Simpson). Adding `tree =` switches
the same call from neutral to phylogenetic diversity.

```{r alpha}
neu <- as_df(hilldiv(bat_counts, q = c(0, 1, 2)))
phy <- as_df(hilldiv(bat_counts, q = c(0, 1, 2), tree = bat_tree))
neu$flavour <- "Neutral (qD)"
phy$flavour <- "Phylogenetic (qPD)"
alpha <- rbind(neu, phy)
alpha$habitat <- hab(alpha$sample)
alpha$flavour <- factor(alpha$flavour, c("Neutral (qD)", "Phylogenetic (qPD)"))

aggregate(value ~ flavour + q + habitat, alpha, function(x) round(mean(x), 2))
```

```{r alpha-plot, fig.alt = "Neutral and phylogenetic alpha diversity by habitat across q"}
ggplot(alpha, aes(factor(q), value, fill = habitat)) +
  geom_boxplot(outlier.size = 0.6, width = 0.7) +
  facet_wrap(~ flavour, scales = "free_y") +
  scale_fill_manual(values = pal, name = NULL) +
  labs(x = "Diversity order (q)", y = "Effective number of prey lineages")
```

Forest specialists and farmland generalists differ only modestly in prey
*richness* (`q = 0`), but the gap widens sharply at higher `q`: at `q = 2` the
forest diet collapses to an effective ~6 prey ASVs against ~12 in farmland,
because a few dominant moths carry most of the forest reads. The phylogenetic
numbers show that the farmland diet is also consistently *broader across the
insect phylogeny* — the generalists span more of the prey tree of life even
where their taxon counts are similar.

## Diversity profiles make the dominance structure explicit

Reading diversity at a single `q` can mislead; a **diversity profile**,
`hillprof()`, traces `qD` continuously and exposes the whole evenness structure
in one curve.

```{r profile, fig.width = 5, fig.height = 3.6, fig.alt = "Neutral diversity profiles, one curve per bat, coloured by habitat"}
prof <- as_df(hillprof(bat_counts, q = seq(0, 3, by = 0.1)))
prof$habitat <- hab(prof$sample)
prof_mean <- aggregate(value ~ q + habitat, prof, mean)

ggplot(prof, aes(q, value, group = sample, colour = habitat)) +
  geom_line(alpha = 0.25, linewidth = 0.3) +
  geom_line(data = prof_mean, aes(group = habitat), linewidth = 1.3) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(x = "Diversity order (q)", y = "Neutral diversity (qD)")
```

The forest profiles fall away more steeply and sit lower — the signature of a
few dominant prey — while the farmland profiles decline gently from a higher
richness, the signature of a more even diet. `hilleven()` condenses this into a
single number:

```{r evenness}
even <- as_df(hilleven(bat_counts, q = 2))
even$habitat <- hab(even$sample)
aggregate(value ~ habitat, even, function(x) round(mean(x), 3))
```

The mean `q = 2` evenness is far lower in forest than in farmland, confirming
that forest diets are dominated by a handful of moth ASVs.

## At which scale does the diet turn over?

A nested design asks a question a single alpha/beta split cannot answer: does
diet vary mostly *between individual bats*, *between roosts*, or *between
habitats*? `hillpart()` answers it directly with a one-sided `hierarchy`
formula, decomposing the pooled (gamma) diversity into one beta **per
hierarchical level**. Each beta is the turnover *gained* at that scale, and the
chain telescopes exactly (`gamma = alpha_finest × ∏ beta`).

```{r hier}
hier_n <- hillpart(bat_counts, q = c(0, 1, 2),
                   hierarchy = ~ habitat / roost, metadata = bat_metadata)
hier_p <- hillpart(bat_counts, q = c(0, 1, 2), tree = bat_tree,
                   hierarchy = ~ habitat / roost, metadata = bat_metadata)
hier_p
```

```{r hier-plot, fig.alt = "Turnover gained at each nested scale, neutral versus phylogenetic"}
beta_n <- as_df(hier_n); beta_p <- as_df(hier_p)
beta_n <- beta_n[beta_n$scale != "sample", ]; beta_n$flavour <- "Neutral"
beta_p <- beta_p[beta_p$scale != "sample", ]; beta_p$flavour <- "Phylogenetic"
betas <- rbind(beta_n, beta_p)
betas$scale <- factor(betas$scale, c("roost", "habitat", "total"),
                      labels = c("among individuals\n(within roost)",
                                 "among roosts\n(within habitat)",
                                 "among habitats"))

ggplot(betas, aes(scale, beta, fill = factor(q))) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  facet_wrap(~ flavour) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  scale_fill_brewer(palette = "Blues", name = "q") +
  labs(x = NULL, y = expression("Turnover gained at scale (" * beta[k] * ")"))
```

The two flavours localise turnover to different scales. Neutral turnover is
split between *individuals* and *habitats*: bats sample different exact prey
ASVs from one mouthful to the next, and the two habitats offer different prey.
But **phylogenetic** turnover is concentrated almost entirely *between habitats*
and is negligible within them: individuals in the same habitat eat
phylogenetically interchangeable prey, while the forest and farmland diets are
drawn from genuinely different parts of the insect tree. This
neutral-vs-phylogenetic, scale-resolved contrast — available across all three
diversity types from one engine — is unique to `hilldiv3`.

## Summary

From a single ASV table and a prey tree, `hilldiv3` delivered: per-sample
neutral and phylogenetic Hill numbers (`hilldiv()`), full diversity profiles
(`hillprof()`) and evenness (`hilleven()`), and a scale-resolved nested
partition of turnover under both flavours (`hillpart()` with a `hierarchy`
formula) — a complete dietary-diversity analysis through one consistent
interface. The neutral-versus-phylogenetic contrast, resolved across the whole
diversity profile *and* across every level of the nested design, is what lets a
single study say not just *how much* diets differ but *in what currency* and *at
which scale*.

## References

* Hill, M.O. (1973). Diversity and evenness. *Ecology*, 54, 427–432.
* Jost, L. (2007). Partitioning diversity into independent alpha and beta
  components. *Ecology*, 88, 2427–2439.
* Alberdi, A. & Gilbert, M.T.P. (2019). A guide to the application of Hill
  numbers to DNA-based diversity analyses. *Mol. Ecol. Resour.*, 19, 804–817.
