---
title: "Use case: three faces of diversity in a gut microbiome"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 150,
  message = FALSE
)
```

```{r setup}
library(hilldiv3)
library(ggplot2)

theme_set(theme_bw(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  legend.position = "top"))

# hilldiv3 results are long-format data.frame subclasses; drop the class so
# base `$<-` / `rbind` (and ggplot2) work without coercion.
as_df <- function(x) { class(x) <- "data.frame"; x }
pal <- c(control = "#1F77B4", treatment = "#D62728")
```

Genome-resolved metagenomics summarises a microbial community as a table of
metagenome-assembled genome (MAG) abundances. Each MAG carries two extra layers
of information that a raw count ignores: its **position on the bacterial
phylogeny** and its **functional attributes** (genome size, GC content, oxygen
tolerance, encoded capabilities). These layers can move independently: an
intervention may reshuffle *which* organisms — and *which lineages* — dominate a
community while leaving *what the community does* untouched, if the incoming
genomes are functionally equivalent to the ones they replace. `hilldiv3`
measures all three flavours from the same count table and the same calls, so
taxonomic, phylogenetic and functional change can be compared on a common,
interpretable scale (effective numbers of lineages) — and, crucially, told
apart.

## The data

We use the bundled simulated data set: **24 MAGs across 12 host gut samples**,
split into a `control` and a `treatment` group of six. The MAGs fall into two
deep bacterial clades, and the intervention **swaps which clade dominates**:
control guts are dominated by clade A, treatment guts by clade B, with taxon
richness left untouched. The two clades are *functional mirrors* of one another
— for every genome in one there is a genome in the other with the same trait
profile — so a swap that is dramatic phylogenetically is invisible functionally.
`gut_tree` is the genome phylogeny and `gut_traits` a trait table mixing
continuous (`genome_size`, `gc_content`), categorical (`oxygen`) and binary
(`motility`) attributes.

```{r data}
group <- rep(c("control", "treatment"), each = 6)
names(group) <- colnames(gut_counts)
metadata <- data.frame(group = group, row.names = colnames(gut_counts))
grp <- function(s) group[s]

dim(gut_counts)                 # 24 MAGs x 12 samples
range(colSums(gut_counts > 0))  # per-sample richness (near-constant)
head(gut_traits)
```

`traits2dist()` turns that mixed trait table into a functional distance with
Gower's coefficient, which handles the different variable types automatically:

```{r fdist}
fdist <- traits2dist(gut_traits)
range(fdist)
```

## Neutral, phylogenetic and functional diversity from one call

The diversity *type* is chosen by what you supply: nothing extra → neutral,
`tree =` → phylogenetic, `dist =` → functional. The interface never changes.

```{r alpha}
neu <- as_df(hilldiv(gut_counts, q = c(0, 1, 2)))
phy <- as_df(hilldiv(gut_counts, q = c(0, 1, 2), tree = gut_tree))
fun <- as_df(hilldiv(gut_counts, q = c(0, 1, 2), dist = fdist))
neu$flavour <- "Neutral (qD)"
phy$flavour <- "Phylogenetic (qPD)"
fun$flavour <- "Functional (qFD)"
alpha <- rbind(neu, phy, fun)
alpha$group   <- grp(alpha$sample)
alpha$flavour <- factor(alpha$flavour,
                        c("Neutral (qD)", "Phylogenetic (qPD)", "Functional (qFD)"))

aggregate(value ~ flavour + q + group, alpha, function(x) round(mean(x), 2))
```

```{r alpha-plot, fig.alt = "Neutral, phylogenetic and functional alpha diversity by group across q"}
ggplot(alpha, aes(factor(q), value, fill = group)) +
  geom_boxplot(outlier.size = 0.6, width = 0.7) +
  facet_wrap(~ flavour, scales = "free_y") +
  scale_fill_manual(values = pal, name = NULL) +
  labs(x = "Diversity order (q)", y = "Effective number of lineages")
```

At the **alpha** (within-sample) level the two groups look almost identical in
every flavour: a single host carries about the same number of effective MAGs,
spread over about the same amount of the phylogeny and the same functional
space, whether or not it received the treatment. Read at the alpha level alone,
the intervention looks like it did nothing. The effect is not in *how much*
diversity each gut holds, but in *which* lineages hold it — a compositional
change that only between-sample analysis can see.

## Where does the community turn over — and where doesn't it?

`hillpart()` partitions diversity between the two groups and reports beta, the
effective number of distinct communities. Running it for each flavour shows
*which* axis the treatment shifts:

```{r partition, fig.width = 5.5, fig.height = 3.6, fig.alt = "Between-group turnover by flavour across q"}
part <- function(type, ...) {
  m <- as_df(hillpart(gut_counts, q = c(0, 1, 2), hierarchy = ~ group,
                      metadata = metadata, ...))
  m <- m[m$scale == "total", c("q", "beta")]
  m$flavour <- type
  m
}
beta <- rbind(part("Neutral"),
              part("Phylogenetic", tree = gut_tree),
              part("Functional",   dist = fdist))
beta$flavour <- factor(beta$flavour, c("Neutral", "Phylogenetic", "Functional"))

ggplot(beta, aes(factor(q), beta, fill = flavour)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  scale_fill_brewer(palette = "Set2", name = NULL) +
  labs(x = "Diversity order (q)",
       y = expression("Between-group turnover (" * beta * ")"))
```

Between-group beta grows strongly with `q` for the **neutral and phylogenetic**
decompositions but stays essentially flat (β ≈ 1) for the **functional** one.
The clade swap replaces the dominant genomes with phylogenetically distant ones,
so taxonomic *and* evolutionary composition turn over almost in lock-step — yet
because the two clades are functional mirrors, the *functional* make-up of the
community barely moves. Distinguishing taxonomic, phylogenetic and functional
turnover on one common beta scale is a core strength of the Hill-number
framework as implemented here, and here it isolates a change that is
phylogenetic but not functional.

## Ordination in three spaces

`hillpair()` accepts the same `tree =` / `dist =` switch, so one workflow
produces ordinations in *neutral*, *phylogenetic* and *functional* space:

```{r ordination, fig.width = 8.5, fig.height = 3.4, fig.alt = "PCoA in neutral, phylogenetic and functional space, coloured by group"}
ord_one <- function(d, lab) {
  pc <- cmdscale(d, k = 2)
  o  <- data.frame(pc, mag = rownames(pc))
  names(o)[1:2] <- c("PCoA1", "PCoA2")
  o$group <- grp(o$mag); o$space <- lab
  o
}
ord <- rbind(
  ord_one(hillpair(gut_counts, q = 1, metric = "C"),                 "Neutral (q = 1)"),
  ord_one(hillpair(gut_counts, q = 1, metric = "C", tree = gut_tree), "Phylogenetic (q = 1)"),
  ord_one(hillpair(gut_counts, q = 1, metric = "C", dist = fdist),   "Functional (q = 1)")
)
ord$space <- factor(ord$space,
                    c("Neutral (q = 1)", "Phylogenetic (q = 1)", "Functional (q = 1)"))

ggplot(ord, aes(PCoA1, PCoA2, colour = group)) +
  geom_point(size = 2.4) +
  stat_ellipse(level = 0.68) +
  facet_wrap(~ space, scales = "free") +
  scale_colour_manual(values = pal, name = NULL) +
  labs(x = "PCoA 1", y = "PCoA 2")
```

The same samples separate cleanly into control and treatment in neutral space
and in phylogenetic space, but collapse onto a single overlapping cloud in
functional space. The picture the three panels paint together — *who* and
*which lineage* differ, *what they do* does not — is one no single distance
could give.

## Functional redundancy

`hillred()` quantifies the **functional redundancy** that underlies this
pattern: it fits the saturating relationship between neutral diversity (number
of genomes) and functional diversity (number of distinct trait profiles) across
samples. A curve that plateaus well below the spread of neutral diversity means
many genomes share functions — so the community can replace genomes without
losing function.

```{r redundancy, fig.width = 5, fig.height = 4, fig.alt = "Per-sample neutral versus functional diversity with fitted saturating curves"}
red <- hillred(gut_counts, q = c(1, 2), dist = fdist)
as_df(red)[, c("q", "redundancy")]
plot(red)
```

The fitted redundancy is high (around 0.6–0.7): functional diversity saturates
well below the spread of neutral diversity, so adding genomes contributes far
less than proportional new function. This is exactly *why* the treatment could
overturn the community's taxonomic and
phylogenetic composition without touching its function — every functional role
lost with clade A is recovered from its mirror in clade B. The redundancy that
`hillred()` measures is the mechanism the beta and ordination analyses revealed.

## Summary

From one MAG table plus a genome tree and a trait table, `hilldiv3` produced a
complete three-flavour analysis through one interface: trait-to-distance
conversion (`traits2dist()`); neutral, phylogenetic and functional alpha
diversity (`hilldiv()`); between-group partitioning for each flavour
(`hillpart()`); ordinations in neutral, phylogenetic and functional space
(`hillpair()`); and functional redundancy (`hillred()`). The unified,
type-switching interface is what lets a single study separate a change that is
taxonomic and phylogenetic from one that is functional — here revealing a
community reorganised in identity but conserved in function.

## References

* Jost, L. (2007). Partitioning diversity into independent alpha and beta
  components. *Ecology*, 88, 2427–2439.
* Chiu, C.-H., Jost, L. & Chao, A. (2014). Phylogenetic beta diversity,
  similarity, and differentiation measures based on Hill numbers. *Ecological
  Monographs*, 84, 21–44.
* Alberdi, A. & Gilbert, M.T.P. (2019). A guide to the application of Hill
  numbers to DNA-based diversity analyses. *Mol. Ecol. Resour.*, 19, 804–817.
```
